MYC overrides HIF-1α to regulate proliferating primary cell metabolism in hypoxia

### 2023-05-16 manuscript/manuscript.qmd
spell checking

Hypoxia requires metabolic adaptations to sustain energetically demanding cellular activities. While the metabolic consequences of hypoxia have been studied extensively in cancer cell models, comparatively little is known about how primary cell metabolism responds to hypoxia. Thus, we developed metabolic flux models for human lung fibroblast and pulmonary artery smooth muscle cells proliferating in hypoxia. Unexpectedly, we found that hypoxia decreased glycolysis despite activation of hypoxia-inducible factor 1α (HIF-1α) and increased glycolytic enzyme expression. While HIF-1α activation in normoxia by prolyl hydroxylase (PHD) inhibition did increase glycolysis, hypoxia blocked this effect. Multi-omic profiling revealed distinct molecular responses to hypoxia and PHD inhibition, and suggested a critical role for MYC in modulating HIF-1α responses to hypoxia. Consistent with this hypothesis, MYC knockdown in hypoxia increased glycolysis and MYC overexpression in normoxia decreased glycolysis stimulated by PHD inhibition. These data suggest that MYC signaling in hypoxia uncouples an increase in HIF-dependent glycolytic gene transcription from glycolytic flux.

@@

To address these questions, we have developed models of bioenergetic carbon flux in human lung fibroblasts (LFs) and pulmonary artery smooth muscle cells (PASMCs) cultured in 21% or 0.5% oxygen. These cells may be exposed to a wide range of oxygen concentrations *in vivo*, continue to proliferate despite hypoxic culture conditions *in vitro*, and play important roles in the pathology of non-cancerous diseases in which tissue hypoxia features prominently, including pulmonary hypertension and pulmonary fibrosis. We found that hypoxia fails to increase glycolysis in these primary cells despite robust up-regulation of the HIF-1α transcriptional program. In normoxia, HIF-1α stabilization by the PHD inhibitor molidustat (BAY-85-3934, "BAY") [@flamme2014] did increase glycolysis and lactate efflux; however, hypoxia blocked this response. These findings suggested that important hypoxia-dependent regulatory mechanisms override the metabolic consequences of HIF-1α-dependent increases in glycolytic gene expression. Transcriptomic profiling identified a critical role for the transcription factor MYC in the adaptive response to hypoxia. Using knockdown and overexpression approaches, we demonstrated that MYC attenuates HIF-driven glycolysis in hypoxia and following HIF stabilization in normoxia.

@@

We observed a similar effect of hypoxia on glycolysis when using RNA interference to silence PHD2 expression and activate HIF-1α gene transcription (**Figure 8 - figure supplement 1**). As with BAY treatment, siPHD2 stabilized HIF-1α in normoxia and increased glycolytic gene expression (**Figure 8 - figure supplement 1B-E**); however, hypoxia inhibited siPHD2-dependent increaes in glycolysis (**Figure 8 - figure supplement 1F**). These data indicate that the metabolic consequences of BAY treatment are a direct consequence of PHD inhibition and not simply an off-target effect.

@@

To examine the role of hypoxia-induced MYC activation in the metabolic response to hypoxia in proliferating primary cells, we first measured MYC by immunoblot. Consistent with our bioinformatic results, we observed increased MYC in hypoxia-treated cells, but not with BAY-treatment alone, where MYC was decreased (**Figure 11A-B**). These findings were similar in siPHD2-treated cells (**Figure 11 - figure supplement 1**). Together, these data suggest a model whereby hypoxia activates MYC and inhibits HIF-driven increases in glycolysis (**Figure 11C, top**). Inhibiting MYC in hypoxia, therefore, should increase glycolysis. Conversely, HIF-dependent glycolysis after BAY treatment in normoxia should be sensitive to inhibition by MYC overexpression (**Figure 11C, bottom**).

@@

We next performed the complementary experiment to determine whether MYC overexpression could attenuate the increase in glycolysis observed with BAY treatment in normoxia (**Figure 11C, bottom**). We did not observe any significant differences in cell growth rate with MYC overexpression (**Figure 11H**). Interestingly, MYC overexpression potentiated BAY-stimulated increases in HIF-1α stabilization and target gene expression (**Figure 11 - figure supplement 2F-J**). In spite of these transcriptional changes, MYC overexpression attenuated BAY-stimulated lactate efflux in normoxia. Together, these data indicate that hypoxia-induced MYC expression may be one factor that uncouples the HIF transcriptional program from glycolytic flux in proliferating primary cells.

@@

Our data indicate that hypoxia-induced MYC expression is one such regulatory mechanism. MYC is a transcription factor that regulates the expression of numerous genes involved in many biological processes, including metabolism, proliferation, apoptosis, and differentiation [@dang2006; @li2020; @stine2015]. As deregulated MYC activity has been associated with the majority of human cancers [@vita2006], much of our understanding of MYC regulation comes from studies using cancer cell models [@li2020; @stine2015; @dang2012; @madden2021]. The role of MYC in the biology of untransformed cells is less well understood. The literature describes a complex and reciprocal relationship between HIF and MYC that depends on both environmental (*e.g.*, hypoxia) and cellular context [@li2020]. Generally, HIF-1 has been observed to inhibit MYC through multiple mechanisms [@koshiji2004; @koshiji2005; @gordan2007; @zhang2007], and this previous work is consistent with our present observations that HIF stabilization following BAY treatment in normoxia decreased MYC protein and target gene expression. Conversely, MYC has been implicated in increased HIF activity through transcriptional and post-transcriptional mechanisms [@zhang2009; @chen2013; @doe2012], primarily in the context of malignant transformation. The observation that MYC may antagonize the transcriptional effects of HIF to sustain primariy cell proliferation and metabolism in hypoxia suggests a substantially different regulatory relationship than has been previously described.

Understanding how MYC transcriptional activity affects hypoxic primary cell metabolism is imperative to our understanding of cellular adaptation to hypoxia. MYC stimulates the expression of nuclear-encoded mitochondrial genes and promotes mitochondrial biogenesis, both directly and through activation of mitochondrial transcriptional factor A (TFAM) [@li2005]. Indeed, we found that the oxidative phosphorylation gene set was enriched in BAY-treated cells cultured in hypoxia (**FIgure 10C**). In this way, hypoxic MYC activation may sustain energy production by oxidative phosphorylation, thereby decreasing the energetic demands that would otherwise drive increased glycolytic flux. Beyond oxidative phosphorylation, MYC regulates genes involved in many other intermediary metabolic pathways, including amino acids, nucleotides, and lipids [@stine2015], that may also impact the central pathways of carbon metabolism studied in this work.

@@

>>>>

Hypoxia requires metabolic adaptations to sustain energetically demanding cellular activities. While the metabolic consequences of hypoxia have been studied extensively in cancer cell models, comparatively little is known about how primary cell metabolism responds to hypoxia. Thus, we developed metabolic flux models for human lung fibroblast and pulmonary artery smooth muscle cells proliferating in hypoxia. Unexpectedly, we found that hypoxia decreased glycolysis despite activation of hypoxia-inducible factor 1α (HIF-1α) and increased glycolytic enzyme expression. While HIF-1α activation in normoxia by prolyl hydroxylase (PHD) inhibition did increase glycolysis, hypoxia blocked this effect. Multi-omic profiling revealed distinct molecular responses to hypoxia and PHD inhibition, and suggested a critical role for MYC in modulating HIF-1α responses to hypoxia. Consistent with this hypothesis, MYC knockdown in hypoxia increased glycolysis and MYC over-expression in normoxia decreased glycolysis stimulated by PHD inhibition. These data suggest that MYC signaling in hypoxia uncouples an increase in HIF-dependent glycolytic gene transcription from glycolytic flux.

@@

To address these questions, we have developed models of bioenergetic carbon flux in human lung fibroblasts (LFs) and pulmonary artery smooth muscle cells (PASMCs) cultured in 21% or 0.5% oxygen. These cells may be exposed to a wide range of oxygen concentrations *in vivo*, continue to proliferate despite hypoxic culture conditions *in vitro*, and play important roles in the pathology of non-cancerous diseases in which tissue hypoxia features prominently, including pulmonary hypertension and pulmonary fibrosis. We found that hypoxia fails to increase glycolysis in these primary cells despite robust up-regulation of the HIF-1α transcriptional program. In normoxia, HIF-1α stabilization by the PHD inhibitor molidustat (BAY-85-3934, "BAY") [@flamme2014] did increase glycolysis and lactate efflux; however, hypoxia blocked this response. These findings suggested that important hypoxia-dependent regulatory mechanisms override the metabolic consequences of HIF-1α-dependent increases in glycolytic gene expression. Transcriptomic profiling identified a critical role for the transcription factor MYC in the adaptive response to hypoxia. Using knockdown and over-expression approaches, we demonstrated that MYC attenuates HIF-driven glycolysis in hypoxia and following HIF stabilization in normoxia.

@@

We observed a similar effect of hypoxia on glycolysis when using RNA interference to silence PHD2 expression and activate HIF-1α gene transcription (**Figure 8 - figure supplement 1**). As with BAY treatment, siPHD2 stabilized HIF-1α in normoxia and increased glycolytic gene expression (**Figure 8 - figure supplement 1B-E**); however, hypoxia inhibited siPHD2-dependent increases in glycolysis (**Figure 8 - figure supplement 1F**). These data indicate that the metabolic consequences of BAY treatment are a direct consequence of PHD inhibition and not simply an off-target effect.

@@

To examine the role of hypoxia-induced MYC activation in the metabolic response to hypoxia in proliferating primary cells, we first measured MYC by immunoblot. Consistent with our bioinformatic results, we observed increased MYC in hypoxia-treated cells, but not with BAY-treatment alone, where MYC was decreased (**Figure 11A-B**). These findings were similar in siPHD2-treated cells (**Figure 11 - figure supplement 1**). Together, these data suggest a model whereby hypoxia activates MYC and inhibits HIF-driven increases in glycolysis (**Figure 11C, top**). Inhibiting MYC in hypoxia, therefore, should increase glycolysis. Conversely, HIF-dependent glycolysis after BAY treatment in normoxia should be sensitive to inhibition by MYC over-expression (**Figure 11C, bottom**).

@@

We next performed the complementary experiment to determine whether MYC over-expression could attenuate the increase in glycolysis observed with BAY treatment in normoxia (**Figure 11C, bottom**). We did not observe any significant differences in cell growth rate with MYC over-expression (**Figure 11H**). Interestingly, MYC over-expression potentiated BAY-stimulated increases in HIF-1α stabilization and target gene expression (**Figure 11 - figure supplement 2F-J**). In spite of these transcriptional changes, MYC over-expression attenuated BAY-stimulated lactate efflux in normoxia. Together, these data indicate that hypoxia-induced MYC expression may be one factor that uncouples the HIF transcriptional program from glycolytic flux in proliferating primary cells.

@@

Our data indicate that hypoxia-induced MYC expression is one such regulatory mechanism. MYC is a transcription factor that regulates the expression of numerous genes involved in many biological processes, including metabolism, proliferation, apoptosis, and differentiation [@dang2006; @li2020; @stine2015]. As deregulated MYC activity has been associated with the majority of human cancers [@vita2006], much of our understanding of MYC regulation comes from studies using cancer cell models [@li2020; @stine2015; @dang2012; @madden2021]. The role of MYC in the biology of untransformed cells is less well understood. The literature describes a complex and reciprocal relationship between HIF and MYC that depends on both environmental (*e.g.*, hypoxia) and cellular context [@li2020]. Generally, HIF-1 has been observed to inhibit MYC through multiple mechanisms [@koshiji2004; @koshiji2005; @gordan2007; @zhang2007], and this previous work is consistent with our present observations that HIF stabilization following BAY treatment in normoxia decreased MYC protein and target gene expression. Conversely, MYC has been implicated in increased HIF activity through transcriptional and post-transcriptional mechanisms [@zhang2009; @chen2013; @doe2012], primarily in the context of malignant transformation. The observation that MYC may antagonize the transcriptional effects of HIF to sustain primary cell proliferation and metabolism in hypoxia suggests a substantially different regulatory relationship than has been previously described.

Understanding how MYC transcriptional activity affects hypoxic primary cell metabolism is imperative to our understanding of cellular adaptation to hypoxia. MYC stimulates the expression of nuclear-encoded mitochondrial genes and promotes mitochondrial biogenesis, both directly and through activation of mitochondrial transcriptional factor A (TFAM) [@li2005]. Indeed, we found that the oxidative phosphorylation gene set was enriched in BAY-treated cells cultured in hypoxia (**Figure 10C**). In this way, hypoxic MYC activation may sustain energy production by oxidative phosphorylation, thereby decreasing the energetic demands that would otherwise drive increased glycolytic flux. Beyond oxidative phosphorylation, MYC regulates genes involved in many other intermediary metabolic pathways, including amino acids, nucleotides, and lipids [@stine2015], that may also impact the central pathways of carbon metabolism studied in this work.

@@

# Source data files

**Figure 1 - source data 1** Uncropped blot images for Figure 1.

**Figure 1 - source data 2** Uncropped blot images for Figure 1 - figure supplement 3.

**Figure 1 - source data 3** Uncropped blot images for Figure 1 - figure supplement 4.

**Figure 2 - source data 1** Uncropped blot images for Figure 2.

**Figure 8 - source data 1** Uncropped blot images for Figure 8 - figure supplement 1.

**Figure 11 - source data 1** Uncropped blot images for Figure 11A.

**Figure 11 - source data 2** Uncropped blot images for Figures 11D and 11G.

**Figure 11 - source data 3** Uncropped blot images for Figure 11 - figure supplement 1.

**Figure 11 - source data 4** Uncropped blot images for Figure 11 - figure supplement 2.
